\documentclass[a4paper, 12pt, openright, oneside, final]{book}
\usepackage{eso-pic}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[english]{babel}
\usepackage{listings}
\usepackage{varioref}
\usepackage{subfigure}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage[pdftex,bookmarks=true,hypertexnames=true]{hyperref}
\usepackage{fancyhdr}
\newcommand{\fncyblank}{\fancyhf{}}
\newenvironment{abstract}%
{\cleardoublepage\fncyblank\null\vfill\begin{center}%
\bfseries\abstractname\end{center}}%
{\vfill\null}

%\lstset{%
%  basicstyle=\small,
%  frame=tb,
%  captionpos=b,
%  stringstyle=\ttfamily,
%  showstringspaces=false,
%  numbers=left,
%  numberstyle=\tiny,
%  framextopmargin=2pt,
%  framexbottommargin=2pt,
%  stepnumber=1,
%  numbersep=5pt}

\usepackage{courier}
\usepackage{color}
\usepackage{xcolor}
\lstset{
         basicstyle=\scriptsize\ttfamily, % Standardschrift
         %numbers=left,               % Ort der Zeilennummern
         numberstyle=\tiny,          % Stil der Zeilennummern
         %stepnumber=2,               % Abstand zwischen den Zeilennummern
         numbersep=5pt,              % Abstand der Nummern zum Text
         tabsize=2,                  % Groesse von Tabs
         extendedchars=true,         %
         breaklines=true,            % Zeilen werden Umgebrochen
         keywordstyle=\color{red},
         frame=b,         
 %        keywordstyle=[1]\textbf,    % Stil der Keywords
 %        keywordstyle=[2]\textbf,    %
 %        keywordstyle=[3]\textbf,    %
 %        keywordstyle=[4]\textbf,   \sqrt{\sqrt{}} %
         stringstyle=\color{black}\ttfamily, % Farbe der String
         showspaces=false,           % Leerzeichen anzeigen ?
         showtabs=false,             % Tabs anzeigen ?
         xleftmargin=5pt,
         framexleftmargin=5pt,
         framexrightmargin=5pt,
         framexbottommargin=4pt,
         %backgroundcolor=\color{lightgray},
         showstringspaces=false      % Leerzeichen in Strings anzeigen ?        
}
\lstloadlanguages{% Check Dokumentation for further languages ...
         %[Visual]Basic
         %Pascal
         C
         %C++
         %XML
         %HTML
         %Java
}
%\DeclareCaptionFont{blue}{\color{blue}} 

%\captionsetup[lstlisting]{singlelinecheck=false, labelfont={blue},
%textfont={blue}}
\usepackage{caption} \DeclareCaptionFont{white}{\color{white}}
\DeclareCaptionFormat{listing}{\colorbox[cmyk]{0.43, 0.35,
0.35,0.01}{\parbox{\textwidth}{\hspace{15pt}#1#2#3}}}
\captionsetup[lstlisting]{format=listing,labelfont=white,textfont=white,
singlelinecheck=false, margin=0pt, font={bf,footnotesize}}


\hypersetup{
pdfauthor= {Fabio Falzoi},
pdftitle= {Design, testing and performance analisys of efficient lock-free solutions 
for multi-core Linux scheduler},
pdfsubject = {Design, testing and performance analisys of efficient 
lock-free solutions for multi-core Linux scheduler },
pdfkeywords = {realtime, scheduling, linux, edf, multiprocessor, multicore, deadline},
pdfborder = { 0 0 0 0 }
}

\onehalfspacing
\begin{document}
\frontmatter

% Frontpage
\input{frontpage}

% dedication
\input{dedication}

% abstract
\begin{abstract}
  Multiprocessor systems are now the de facto preferred computing
  platform for many application domains, including personal computers
  and server workstations.
  
  The benefits of multi-core technology in terms of increased
  computational power with a reduced energy consumption, are now being
  used for more implementing efficient embedded devices and personal
  appliances like smart phones and tablets. 
  
  A popular OS like Linux, which was not originally designed to be a
  Real-Time Operating System (RTOS), is now being used for embedded
  real-time systems with multi-core platforms. Consequently, many
  flavors of Linux now include a real-time scheduler. One recent
  example of real-time scheduler for Linux is the SCHED\_DEADLINE
  patch, an implementation of the popular ``Earliest Deadline First''
  algorithm. Such scheduler can be useful also for large many-core
  server workstations, because it helps to control the quality of
  service of the user requests. For this reason, it is important that
  the scheduler implementation to be efficient and scale well with the
  number of processors.

  In this thesis, I present two original contributions to the area of
  real-time scheduling in the Linux kernel. First, I present PRACTISE,
  a tool to develop, debug, test and analyse real-time scheduling data
  structures in user space. Unlike other similar tools, PRACTISE
  executes code in parallel, allowing to test and analyse the
  performance of the code in a realistic multiprocessor scenario.

  Then, I present several data structures for implementing a
  distributed queue to efficiently support global scheduling in a
  large multi-core: max-heap and min-heap, skip-list with a
  flat-combiner strategy, and a novel algorithm called
  \emph{fast-cache}. I compare the different data structures and
  algorithms using both PRACTISE and directly in the kernel.
\end{abstract}


\tableofcontents
\listoffigures
%\listoftables

% Introduction to the work
\input{introduction}

\mainmatter
% background
\input{background}

\input{synchro_mech_analysis}

\input{data_structures_developed}

\input{practise}

% results
\input{exp-results}

% conclusions
\input{conclusions}

% code listings, etc.
\input{appendix}

\backmatter

\input{acknowledgments}

\bibliography{bibliography}
\bibliographystyle{plain}
\end{document}
